### 2017-11-08 RefReports/replyToReferees.tex

\newpage
\section*{Reply to Associate Editor}

\re{
Associate Editor Comments:
\\
In addition to reviewers with respect to meeting scientific standards for reporting, it seems this work currently does not meet the MIFlowCyt requirements (not provided, as required by Cytometry A), especially as noted  with respect to data availability. Also missing are details of available implementations of their approach, per  Cytometry A software and data sharing requirements.}

\smallskip
Dear Associate Editor, we now make publicly available all data (including MIFlowCyt checklists) and analysis with the manuscript, organized as described below. We have added to the \emph{Supplemental information} the description below.

\begin{itemize}
\item All MATLAB scripts used to generate the plots and analyze the data in
\\
{\small \href{https://github.com/AmirErez/BimodalLogspaceCytA/tree/master/Scripts}{https://github.com/AmirErez/BimodalLogspaceCytA/tree/master/Scripts}}
\item All data used in the manuscript and accompanying MIFlowCyt checklists, which comes from two datasets:
	\begin{enumerate}
	\item 20140920-OT1-dynamics - used for Fig. 3 in the manuscript.
	\item FeinermanScience2008 - previously published in Feinerman et al. Science Vol. 321, Issue 5892, pp. 1081-1084 (2008). 
	\end{enumerate}
These data with their respective MIFlowCyt checklists are freely available in

@@

\smallskip
We thank the Referee for pointing out this lack of focus in the methods section. The Referee is correct in understanding that some of our purpose here is to show that determining the number of peaks of experimental data is a difficult task. A second purpose is to refer to recent literature which employs Hartigan's test for a similar experimental situation. This is the relevance and particular choice of Hartigan as an example. Taking into account the Referee's comment, we have shortened the discussion of Hartigan's method in the main text, moving its bulk to the supplementary section for reference. We hope that in this way we clarify the focus of the methods section as per the Referee's suggestion.

\re{
4) In that section, the authors assume that $log(I)$ is a Gaussian mixture model. Tedious, unenlightening computations are then displayed much to this reader's displeasure, where only the outcome is necessary to be relayed. This is made more egregious due to the use of S (normally reserved for a survival function, i.e.  one minus the cdf, in all stats papers) and F (normally reserved for a cdf).  Essentially much of pages 4 and 5 could be deleted on the assumption that a sufficiently interested reader probably knows how to take derivatives and/or work mathematica, and the chase can be cut to.
}

\smallskip
We accept the Referee's comment that the derivations do not need to be spelled out in full in the main body of the manuscript. Accordingly, we have moved some to the \emph{Supporting information} section, in our attempt to accommodate Cytometry A's readership which is of heterogeneous mathematical background. We are pleased to accommodate the Referee's suggestions that the letters $S$ and $F$ be left alone, and instead have changed $F\rightarrow A$ and $S_{1,3}\rightarrow B_{1,3}$, which we hope the Referee will find less objectionable.

@@

\smallskip
We hope the Referee understands that the same logic that guided our choice of $P,Q$ has led us to choose $P_2$, namely: (i) it is consistent with $P(\log I)$ in that $P$ is used to describe the pdf of the logarithm of intensity; (ii) $P_2$, instead of $P$, is used to make explicit the fact that these functions are different, not just of a different argument; (iii) $P$ is consistent with this journal's accepted abbreviations which specify $P$ for probability.

\smallskip
We have added a short comment about the multivariate change of variables, which lies outside the focus of the work. In our analysis, we focus on the marginal distributions which are univariate. We fear a deeper discussion of this will distract the readers from our main message.

\re{
8) From that section on, what was used in the present notation as a value a pdf is evaluated at, e.q. $Q(I)$, becomes a measured value.  Moreover the conditioning etc. is all written in a non-formal, non-standard way. Such abuse of notation is alleviated with use of standard notation, which would be advised.}

\smallskip
We believe that we have used standard notation. We refer the Referee to the book by Koller and Friedman (ref. 32 in the main text). Prof. Koller's book is widely acclaimed in the computational biology field, a highly cited standard reference (according to Google Scholar, 4438 citations since its publication in year 2009). The notation in the conditioning part of the manuscript matches that in Prof. Koller's book. The field of computational biology combines several disciplines with different notation conventions.

### 2017-11-17 Manuscript/LogSpaceDistributionsCytometryA.tex
Changed figures to tif
tif
Update

\begin{flushleft}
{\Large
\textbf\newline{Modeling of cytometry data in logarithmic space: when is a bimodal distribution not bimodal?} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Amir Erez\textsuperscript{1*},
Robert Vogel\textsuperscript{2},
Andrew Mugler\textsuperscript{3},
Andrew Belmonte\textsuperscript{1,4},
Gr\'egoire Altan-Bonnet\textsuperscript{1}

\bigskip
\textbf{1} Immunodynamics Group, Cancer and Inflammation Program, National Cancer Institute, National Institutes of Health, Bethesda, Maryland 20814, USA
\\
\textbf{2} IBM T. J. Watson Research Center, Yorktown Heights, New York 10598, USA
\\
\textbf{3} Department of Physics and Astronomy, Purdue University, West Lafayette, Indiana 47907, USA
\\
\textbf{4} Department of Mathematics, Pennsylvania State University, University Park, Pennsylvania 16802, USA
\bigskip

% Use the asterisk to denote corresponding authorship and provide email address in note below.
* erezam@bgu.ac.il
\end{flushleft}

@@

\begin{abstract}
Recent efforts in systems immunology lead researchers to build quantitative models of cell activation and differentiation. One goal is to account for the distributions of proteins from single-cell measurements by flow cytometry or mass cytometry as a readout of biological regulation. In that context, large cell-to-cell variability is often observed in biological quantities. We show here that these readouts, viewed in logarithmic scale may result in two easily-distinguishable modes, while the underlying distribution (in linear scale) is unimodal. We introduce a simple mathematical test to highlight this mismatch. We then dissect the flow of influence of cell-to-cell variability using a graphical model and its effect on measurement noise. Finally we show how acquiring additional biological information can be used to reduce uncertainty introduced by cell-to-cell variability, helping to clarify whether the data is uni- or bimodal. 
This communication has cautionary implications for manual and automatic gating strategies, as well as clustering and modeling of single-cell measurements. 
\end{abstract}

@@

\section*{Methods - theory}
Given the pdf $P(\log I)$, one can formally derive $Q(I)$ as \cite{Mukhopadhyay2000},
\bea 
\label{eq:P2Q}
Q(I) &=& P(\log I) \left\vert \frac{d}{dI} \log I \right\vert \nn
 &=& \frac{1}{I}P(\log I) = e^{-y} P(y)\,,
\eea
with $\log I\equiv y$. This relation serves us both in this theoretical section and in the experimental \emph{Results} section, where we estimate the empirical distribution for $Q(I)$ using this relation. We now address the question: when is $Q(I)$ unimodal while $P({\log I})$ is multimodal ?
\smallskip

@@

We define $y=\log I$ and proceed to find the number of extrema for $P(y)$ and $Q(I)$. We do this by recasting Eq.~\ref{eq:LogNormalMix} and its derivatives into the functions $A(y),B_1(y),B_3(y)$. Full mathematical details can be found in the \emph{Supporting information}. We introduce the following functions,
\bea 
\label{eq:extrema}
A(y) &=& \left(\frac{\alpha}{1-\alpha}\right)\left(\frac{\sigma_1}{\sigma_2}\right) e^{\frac{(y-y_1)^2}{2 \sigma_1^2}-\frac{(y-y_2)^2}{2 \sigma_2^2}}  \nn
B_1(y) &=& \frac{\frac{y-y_1}{\sigma_1^2}+1}{\frac{y_2-y}{\sigma_2^2}-1} \nn
B_3(y) &=& \frac{(y-y_1)/\sigma_1^2}{(y_2-y)/\sigma_2^2}\,.
\eea
To find the extrema of $Q(I)$, one must find the solutions $y^*$ for $B_1(y^*)=A(y^*)$. Similarly, for the extrema of $P(\log I)$ one must solve $B_3(y^*)=A(y^*)$.

\smallskip
$A(y)$ is the ratio of the two Gaussians in Eq.~\ref{eq:LogNormalMix} and is therefore always non-negative; this implies that any extremum $y_*$ must satisfy $B_1(y_*)\ge 0$ and $B_3(y_*)\ge 0$. We note that the condition for extrema in $Q(I)$ requires that $\frac{d}{dy}P(y)=P(y)$ whereas, of course, extremizing $P(\log I)$ sets its derivative to zero, demonstrating the fact that the loci of the modes for $P(\log I)$ and $Q(I)$ are manifestly different. The region where the log-space distribution shows a second mode occurs when for $B_3(y^*)=A(y^*)$ admits three solutions whereas $B_1(y^*)=A(y^*)$ admits only one solution. Given that these equations are transcendental, a graphical way to asses the number of solutions is to plot $A(y),B_1(y),B_3(y)$ and count the number of times $B_1$ and $B_3$ intersect $A$.\smallskip

@@

\begin{figure}[!ht]
 \centering
  \includegraphics[clip,width=1\hsize]{Fig2.png}
  \caption{\textbf{Graphical solution to count the number of extrema.} When the red (blue) curves intersect the dashed black line, $P(\log I)$ ($Q(I)$) are extremized. Dashed black: $A(y)$, blue: $B_1(y)$ and red: $B_3(y)$. The mismatch between the number of extrema of $P(\log I)$ and $Q(I)$ is apparent when the red curve intersects $A$ at 3 points, whereas the blue curve only intersects $A$ once. In both cases, the loci of the extrema are different for the two distributions. The plots along the diagonal (which correspond to the cases in Fig.~\ref{fig:PlotsLinearLog}) show the case $\sigma_1=\sigma_2$ which simplifies $A(y)$ to a straight line (the general case being a parabola) in these axes.}
  \label{fig:ExtremaSolutions}
\end{figure}

In the plots along the diagonal, we have $\sigma_1=\sigma_2$ (as in Fig.~\ref{fig:PlotsLinearLog})  which simplifies $A(y)$ since the quadratic (Gaussian) terms cancel, leaving only an exponential. This leads to a simple criterion to determine whether $P(\log I)$ itself admits one or two modes - previously in Fig.~\ref{fig:PlotsLinearLog}(right) we saw an example where $P(\log I)$ is unimodal despite being generated from a mixture. Graphically, we see that for $B_3=F$ to have 3 solutions, $\log B_3(y)$ has to have a slope less than $\log A(y)$ about the extremum $y_*$. In other words, $\left.\frac{d}{dy} \log B_3(y)\right\vert_{y_*} \le \left.\frac{d}{dy} \log A(y)\right\vert_{y_*}$, with equality as the threshold between 1 and 3 extrema. This leads to the following intuitive criterion,
\be 
\label{eq:Plogy3Maxima}
(y_* - y_1)(y_2 - y_*) \ge \sigma ^2 \,\,\Longrightarrow \,\, \mbox{3 extrema for } P(\log I)\,,
\ee
which states that for $P(\log I)$ to appear bimodal, it must have an extremum ($y_*$) such that the variance of the individual Gaussian components of $P(\log I)$ must be smaller than the distance between $y_*$ and the Gaussian centers. Substituting for $y_* \approx \log 316$, $y_1 = \log 100$ and $y_2=\log 1000$ and $\sigma^2 = 1.44$ we see that the criterion in Eq.~\ref{eq:Plogy3Maxima} is not satisfied and indeed in Fig.~\ref{fig:PlotsLinearLog}(right) and Fig.~\ref{fig:ExtremaSolutions}(bottom right) we see that $P(\log I)$ has only one mode. A similar condition can be derived for $Q(I)$, that is, $\left.\frac{d}{dy} \log B_1(y)\right\vert_{y_*} \le \left.\frac{d}{dy} \log A(y)\right\vert_{y_*}$, such that,
\be 
\label{eq:QI3Maxima}
(y_* - y_1+\sigma^2)(y_2 - y_* - \sigma^2) \ge \sigma ^2\,\,\Longrightarrow \,\,\mbox{3 extrema for } Q(I)\,,
\ee \smallskip
it is, however, hard to compare the two bounds analytically because the $y_*$ which extremizes $P(\log I)$ is different from the $y_*$ which extremizes $Q(I)$.

@@

We now apply our analysis to experimental data, namely the measured distribution of Extracellular Signal-regulated Kinase (ERK) phosphorylation (ppERK) signaling in CD8+ primary mouse T-cells responding to antigens and inhibited by the SRC inhibitor Dasatinib. These data were acquired in exactly the same manner as the experiments in Ref.~\cite{Vogel2016}, for brevity we refer the reader there for all experimental details. In Fig.~\ref{fig:Experiment}(A) we see how a commercially-available analysis software (FlowJo \cite{FlowJo}) plots the distribution of ppERK in such an experiment, which clearly shows a bimodal structure. Fig.~\ref{fig:Experiment}(B) plots those same data, formally $P(\log I) := \hat{f}_{\log I}^{(8751)}$, the probability density estimated from the data by taking the logarithm of the intensity of 8751 cells and binning it in bins of size 0.2. Our procedure for estimating $P(\log I)$ gives the two modes as in FlowJo (red dots) whereas $Q(I)$ (derived using Eq.~\ref{eq:P2Q}) has a single mode (blue dots). We fit $P(\log I)$ as a Gaussian mixture. This is followed in Fig.~\ref{fig:Experiment}(C) by the same extrema analysis as in Fig.~\ref{fig:ExtremaSolutions}, revealing that indeed $Q(I)$ has a single maximum.
\smallskip

\begin{figure}[!ht]
 \centering
  \includegraphics[width=1\linewidth]{Fig3.png}
  \caption{\textbf{Analysis of experimental data reveals the effect we describe in a real scenario. }\textbf{(A)} Histogram of ppERK as plotted by FlowJo \cite{FlowJo}; \textbf{(B)} Histograms for $P(\log I)$ (red,dots) and $Q(I)$ (blue,dots) vs. $\log I$ as estimated from the data binned logarithmically. Note that plotting $Q(I)$ vs. $\log I$ is somewhat unusual but allows both $P$ and $Q$ to be plotted on the same axis. The red line shows the result of fitting $P(\log I)$ to a gaussian mixture model (Eq.~\ref{eq:LogNormalMix}), and the blue line is the estimate for $Q(I)$ from $P(\log I)$ according to Eq.~\ref{eq:P2Q}. The blue star indicates the location of the only maximum for $Q(I)$ obtained from Eq. \ref{eq:extremaQ}, despite the obvious two maxima in $P(\log I)$ (red). Hartigan's unimodality p-values for $\log I$ (red) and $I$ (blue) are taken directly from the data without binning, corroborating that $\log I$ is bimodal whereas $I$ is unimodal. \textbf{(C)} Graphic solution of the extrema conditions as in Fig.~\ref{fig:ExtremaSolutions} explicitly reveals the three solutions for $P(\log I_*)$ (red line intersects black dashed) as opposed to the single solution for $Q(I_*)$ (the blue line intersects the black dashed line up beyond the plotted area, solution also plotted as blue star in the middle plot), indicating that $Q(I)$ has only one mode.}
  \label{fig:Experiment}
\end{figure}
Given the prevalence and success of manual gating of FCM data in many situations, we wondered whether gating in logarithmic scale could be justified \emph{a posteriori}, based on biological knowledge. In the final part of our analysis, we aimed to include additional information such that the biological significance of the distributions in our single-cell measurements is better captured.

\smallskip
We briefly sketch what follows: by considering both ERK1 and ppERK levels, (the two being biologically related and experimentally correlated), we demonstrate that dividing one by the other ameliorates the mismatch between the logarithmic and linear representations. This is done by examining their \emph{joint} probability density function $P_2$; we suggest a probabilistic graphical model that describes the correlations in the system and their dependence on cell-to-cell variability; we suggest a way to normalize for cell-to-cell variability, such that the mismatch (resulting from increased uncertainty) between logarithmic and linear gates is resolved. We then generalize our example and provide a simple experimental and theoretical framework that allows less ambiguous determination of positive/negative gates.

\smallskip
In Fig.~\ref{fig:ppERKvsERK1heatmap}(A), we show the experimental data we will use in our proposed solution. Here we returned to our single-cell measurements of ERK phosphorylation in primary mouse T cells in Ref.~\cite{Feinerman2008}. We show a heat map of the \emph{joint} distribution of ppERK ($I_{ppERK}$) and total ERK1 ($I_{ERK1}$) expression in mouse CD8+ T-cells. We estimate their joint distribution by using a kernel density estimator \cite{botev2010}, which is used only for presentation purposes and for demonstrating their conditional independence (Fig.~\ref{S1_Fig}).
Notably, whereas the two modes of ppERK significantly overlap when plotted in the marginal distribution $P(\log I_{ppERK})$, ppERK expression correlates with total ERK1 levels in their joint distribution $P_2(\log I_{ppERK},\log I_{ERK1})$. The relation in Eq.~\ref{eq:P2Q} can be generalized to $P_2(\log I_{ppERK},\log I_{ERK1})$ (see, eg., \cite{Mukhopadhyay2000}), however, it will not be used in this manuscript.

>>>>

\begin{flushleft}
{\Large
\textbf\newline{Modeling of cytometry data in logarithmic space: when is a bimodal distribution not bimodal?} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Amir Erez\textsuperscript{1*,5},
Robert Vogel\textsuperscript{2},
Andrew Mugler\textsuperscript{3},
Andrew Belmonte\textsuperscript{1,4},
Gr\'egoire Altan-Bonnet\textsuperscript{1}

\bigskip
\textbf{1} Immunodynamics Group, Cancer and Inflammation Program, National Cancer Institute, National Institutes of Health, Bethesda, Maryland 20814, USA
\\
\textbf{2} IBM T. J. Watson Research Center, Yorktown Heights, New York 10598, USA
\\
\textbf{3} Department of Physics and Astronomy, Purdue University, West Lafayette, Indiana 47907, USA
\\
\textbf{4} Department of Mathematics, Pennsylvania State University, University Park, Pennsylvania 16802, USA
\\
\textbf{5} Present address: Department of Molecular Biology, Princeton University, Lewis Thomas Laboratory, Princeton, New Jersey 08544, USA
\bigskip

% Use the asterisk to denote corresponding authorship and provide email address in note below.
* aerez@princeton.edu
\end{flushleft}

@@

\begin{abstract}
Recent efforts in systems immunology lead researchers to build quantitative models of cell activation and differentiation. One goal is to account for the distributions of proteins from single-cell measurements by flow cytometry or mass cytometry as a readout of biological regulation. In that context, large cell-to-cell variability is often observed in biological quantities. We show here that these readouts, viewed in logarithmic scale may result in two easily-distinguishable modes, while the underlying distribution (in linear scale) is unimodal. We introduce a simple mathematical test to highlight this mismatch. We then dissect the flow of influence of cell-to-cell variability proposing a graphical model which motivates higher-dimensional analysis of the data. Finally we show how acquiring additional biological information can be used to reduce uncertainty introduced by cell-to-cell variability, helping to clarify whether the data is uni- or bimodal. 
This communication has cautionary implications for manual and automatic gating strategies, as well as clustering and modeling of single-cell measurements. 
\end{abstract}

@@

\section*{Methods - theory}
We focus this \emph{Methods} section on the theoretical aspect of this work. Brief details regarding the experimental methods can be found at the end of this section.

\smallskip
Given the pdf $P(\log I)$, one can formally derive $Q(I)$ as \cite{Mukhopadhyay2000},
\bea 
\label{eq:P2Q}
Q(I) &=& P(\log I) \left\vert \frac{d}{dI} \log I \right\vert \nn
 &=& \frac{1}{I}P(\log I) = e^{-y} P(y)\,,
\eea
with $\log I\equiv y$. This relation serves us both in this theoretical section and in the experimental \emph{Results} section, where we estimate the empirical distribution for $Q(I)$ using this relation. We now address the question: when is $Q(I)$ unimodal while $P({\log I})$ is multimodal ?
\smallskip

@@

We define $y=\log I$ and proceed to find the number of extrema for $P(y)$ and $Q(I)$. We do this by simple differentiation, demanding that the derivatives of $P$ and $Q$ equal zero as a definition of extrema. This is achieved by recasting Eq.~\ref{eq:LogNormalMix} and its derivatives into the functions $A(y),B_1(y),B_3(y)$, defined as,
\bea 
\label{eq:extrema}
A(y) &=& \left(\frac{\alpha}{1-\alpha}\right)\left(\frac{\sigma_1}{\sigma_2}\right) e^{\frac{(y-y_1)^2}{2 \sigma_1^2}-\frac{(y-y_2)^2}{2 \sigma_2^2}}  \nn
B_1(y) &=& \frac{\frac{y-y_1}{\sigma_1^2}+1}{\frac{y_2-y}{\sigma_2^2}-1} \nn
B_3(y) &=& \frac{(y-y_1)/\sigma_1^2}{(y_2-y)/\sigma_2^2}\,.
\eea
The outcome of this approach is that, to find the extrema of $Q(I)$, one must find the solutions $y_{*q}$ for $B_1(y_{*q})=A(y_{*q})$. Similarly, for the extrema of $P(\log I)$ one must solve $B_3(y_{*p})=A(y_{*p})$. Full mathematical details on how these are derived can be found in the \emph{Supporting information}.

\smallskip
$A(y)$ is the ratio of the two Gaussians in Eq.~\ref{eq:LogNormalMix} and is therefore always non-negative; this implies that any extremum $y_*$ must satisfy $B_1(y_{*q})\ge 0$ and $B_3(y_{*p})\ge 0$. We note that the condition for extrema in $Q(I)$ requires that $\frac{d}{dy}P(y)=P(y)$ whereas, of course, extremizing $P(\log I)$ sets its derivative to zero, demonstrating the fact that the loci of the modes for $P(\log I)$ and $Q(I)$ are manifestly different. The region where the log-space distribution shows a second mode occurs when for $B_3(y_{*p})=A(y_{*p})$ admits three solutions whereas $B_1(y_{*q})=A(y_{*q})$ admits only one solution. Given that these equations are transcendental, a graphical way to asses the number of solutions is to plot $A(y),B_1(y),B_3(y)$ and count the number of times $B_1$ and $B_3$ intersect $A$.\smallskip

@@

\begin{figure}[!ht]
 \centering
  \includegraphics[clip,width=1\hsize]{Fig2.png}
  \caption{\textbf{Graphical solution to count the number of extrema.} We test our peak counting method, based on the same equations as in Fig.~\ref{fig:PlotsLinearLog}, but with additional settings. (Vertical axis represents the functions $A, B_1,B_3$ according to the legend). When the red (blue) curves intersect the dashed black line, $P(\log I)$ ($Q(I)$) are extremized. Dashed black: $A(y)$, blue: $B_1(y)$ and red: $B_3(y)$. The mismatch between the number of extrema of $P(\log I)$ and $Q(I)$ is apparent when the red curve intersects $A$ at 3 points, whereas the blue curve only intersects $A$ once. In both cases, the loci of the extrema are different for the two distributions. The plots along the diagonal (which correspond to the cases in Fig.~\ref{fig:PlotsLinearLog}) show the case $\sigma_1=\sigma_2$ which simplifies $A(y)$ to a straight line (the general case being a parabola) in these axes.}
  \label{fig:ExtremaSolutions}
\end{figure}

In the plots along the diagonal, we have $\sigma_1=\sigma_2$ (as in Fig.~\ref{fig:PlotsLinearLog})  which simplifies $A(y)$ since the quadratic (Gaussian) terms cancel, leaving only an exponential. This leads to a simple criterion to determine whether $P(\log I)$ itself admits one or two modes - previously in Fig.~\ref{fig:PlotsLinearLog}(right) we saw an example where $P(\log I)$ is unimodal despite being generated from a mixture. Graphically, we see that for $B_3=A$ to have 3 solutions, $\log B_3(y)$ has to have a slope less than $\log A(y)$ about the extremum $y_{*p}$. In other words, $\left.\frac{d}{dy} \log B_3(y)\right\vert_{y_{*p}} \le \left.\frac{d}{dy} \log A(y)\right\vert_{y_{*p}}$, with equality as the threshold between 1 and 3 extrema. This leads to the following intuitive criterion,
\be 
\label{eq:Plogy3Maxima}
(y_{*p} - y_1)(y_2 - y_{*p}) \ge \sigma ^2 \,\,\Longrightarrow \,\, \mbox{3 extrema for } P(\log I)\,,
\ee
which states that for $P(\log I)$ to appear bimodal, it must have an extremum ($y_{*p}$) such that the variance of the individual Gaussian components of $P(\log I)$ must be smaller than the distance between $y_{*p}$ and the Gaussian centers. Substituting for $y_{*p} \approx \log 316$, $y_1 = \log 100$ and $y_2=\log 1000$ and $\sigma^2 = 1.44$ we see that the criterion in Eq.~\ref{eq:Plogy3Maxima} is not satisfied and indeed in Fig.~\ref{fig:PlotsLinearLog}(right) and Fig.~\ref{fig:ExtremaSolutions}(bottom right) we see that $P(\log I)$ has only one mode. A similar condition can be derived for $Q(I)$, that is, $\left.\frac{d}{dy} \log B_1(y)\right\vert_{y_{*q}} \le \left.\frac{d}{dy} \log A(y)\right\vert_{y_{*q}}$, such that,
\be 
\label{eq:QI3Maxima}
(y_{*q} - y_1+\sigma^2)(y_2 - y_{*q} - \sigma^2) \ge \sigma ^2\,\,\Longrightarrow \,\,\mbox{3 extrema for } Q(I)\,,
\ee \smallskip
it is, however, hard to compare the two bounds analytically because the $y_{*p}$ which extremizes $P(\log I)$ is different from the $y_{*q}$ which extremizes $Q(I)$.

\subsection*{Experimental methods}
All data and MATLAB scripts used for the experimental part of this work, together with accompanying MIFlowCyt checklists detailing the data analysis, are freely available online, as follows:

\smallskip
\begin{itemize}
\item All MATLAB scripts used to generate the plots and analyze the data in {\small \href{https://github.com/AmirErez/BimodalLogspaceCytA/tree/master/Scripts}{github.com/AmirErez/BimodalLogspaceCytA/tree/master/Scripts}}
\item Al data used in the manuscript and accompanying MIFlowCyt checklists come from two experiments:
	\begin{enumerate}
	\item 20140920-OT1-dynamics - used for Fig. 3 in the manuscript.
	\item FeinermanScience2008 - previously published in Feinerman et al. Science Vol. 321, Issue 5892, pp. 1081-1084 (2008). 
	\end{enumerate}
These data with their respective MIFlowCyt checklists are available in {\small
\href{https://github.com/AmirErez/BimodalLogspaceCytA/tree/master/Data}{github.com/AmirErez/BimodalLogspaceCytA/tree/master/Data}}
\end{itemize} 
All are accessible in {\small\href{https://github.com/AmirErez/BimodalLogspaceCytA}{github.com/AmirErez/BimodalLogspaceCytA}}.
We briefly sketch the methods for each of the two experiments:
\subsubsection*{Experiment 1:} Primary mouse T lymphocytes (OT-1) were activated ex vivo and cultured. OT-1 cells were activated by peptide (SIINFEKL) treated and irradiated (3,000 RAD) primary antigen presenting cells (APCs) from a C57BL/6 mouse. OT-1 cells were labeled with an amine-reactive dye, CTV, according to the manufacture’s protocol (Molecular Probes) for in silico identification. Cells were allowed to rest for one hour after CTV staining, and then distributed in a 96-well v-bottom plate. Cells were then incubated with unique doses of either SRC or MEK inhibitor at a temperature of $37^o$C for 5 minutes. After which,  we added the peptide pulsed APCs (10 APCs to 1 OT-1 T cell) and pelleted mixture of cells by centrifugation for 10 s at 460 rcf at room temperature.  The pellet was then incubated at a temperature of $37^o$C for 10 minutes, followed by 15 minutes of chemical fixing with 2\% PFA on ice and then permeabilization with ice cold 90\% MeOH. Samples were kept at a temperature of $-20^o$C until labeling for FCM.

\subsubsection*{Experiment 2:} C57BL/6N splenocytes were pulsed for 2hr with 100nM SIINFEKL peptide, then irradiated (3000RAD), washed once and used as stimulator/feeder cells. OT-1 lymphocytes were harvested from auxiliary, brachial and inguinal lymph nodes as well as spleen (splenocytes were treated with ACK lysis buffer to remove red blood cells), and mixed with SIINFEKL-pulsed C57BL/6N splenocytes in complete RPMI. After two days, cells were expanded by diluting 2 fold into medium containing 100 pM IL-2. After four days, the cells were again expanded by 2 fold dilution into medium with IL-2. After one more day of culture, cells were harvested and spun through a 1.09 density Ficoll- Paque Plus gradient (GE Healthcare) to remove dead cells. Live cells were recovered, washed twice in complete medium and resuspended at 1 million/ml in complete medium with 100pM IL-2. Cells were used for experiments between 6 and 8 days after primary stimulation. The ppERK response to SRC and MEK inhibition was measured using primary OT-1 T-lymphocytes activated with RMA-S APCs. RMA-S cells were suspended in culture with 1 nM SIINFEKL peptide for 2 h at 37 $^o$C, 5\% CO\textsubscript{2}, and on a rotator to guarantee mixing. During this time we labelled OT-1 cells with an amine-reactive dye, CTV, according to the manufacture’s protocol (Molecular Probes). We rested the OT-1 cells one hour after CTV staining, and then distributed them in a 96-well v-bottom plate. Each well was given various doses of SRC inhibitor and MEK inhibitor and kept at 37 $^o$C for 5 min. Following the 5 min exposure to the inhibitors, we added the peptide pulsed RMA-S (10 RMA-S to 1 OT-1 T cell) and pelleted by centrifugation for 10 s at 460 rcf at room temperature. This step guaranteed that both cell types, OT-1 and RMA-S, came into contact. The cells were allowed to activate for 10 min, followed by fixing on ice in 2\% PFA, and then permeabilized and stored in 90\% MeOH at  -20 $^o$C. APCs were pulsed with serial dilutions of OVA or variant peptides for 2 hr at 37$^o$C, then washed with T cell medium at the time of harvest, and resuspended with anti-CD8α (53-6.72)-Fab-coated T cells in their conditioned media in a V-bottom 96-well plate (Corning). Fab-coating was performed 10 min before cell use by incubating T cells with 10 μg/ml of Fab fragment. T:APC cell contacts were synchronized using a quick centrifugal spin (10s at 400g). Plates containing T:APC conjugates were placed on a water bath at 37$^o$C and incubated for 5 min. Supernatants were then discarded, T:APC conjugates disrupted by vortexing, and cells resuspended in ice-cold 4\% paraformaldehyde for 15 min. Cells were then permeabilized with ice-cold 90\% methanol for 15 min on ice, and washed twice with FACS buffer.

@@

We now apply our analysis to experimental data, by analyzing two different experiments, described briefly in the \emph{Methods} section and in more detail in the MIFlowCyt checklists available online. Both experiments measure the distribution of Extracellular Signal-regulated Kinase (ERK) phosphorylation (ppERK) signaling in CD8+ primary mouse T-cells responding to antigens.

\smallskip
In experiment 1, the T-cells were inhibited by the SRC inhibitor Dasatinib. In Fig.~\ref{fig:Experiment}(A) we see how a commercially-available analysis software (FlowJo \cite{FlowJo}) plots the distribution of ppERK in such an experiment, which clearly shows a bimodal structure. Fig.~\ref{fig:Experiment}(B) plots those same data, formally $P(\log I) := \hat{f}_{\log I}^{(8751)}$, the probability density estimated from the data by taking the logarithm of the intensity of 8751 cells and binning it in bins of size 0.2. Our procedure for estimating $P(\log I)$ gives the two modes as in FlowJo (red dots) whereas $Q(I)$ (derived using Eq.~\ref{eq:P2Q}) has a single mode (blue dots). We fit $P(\log I)$ as a Gaussian mixture. This is followed in Fig.~\ref{fig:Experiment}(C) by the same extrema analysis as in Fig.~\ref{fig:ExtremaSolutions}, revealing that indeed $Q(I)$ has a single maximum.
\smallskip

\begin{figure}[!ht]
 \centering
  \includegraphics[width=1\linewidth]{Fig3.png}
  \caption{\textbf{Analysis of experimental data reveals the effect we describe in a real scenario. }\textbf{(A)} Histogram of ppERK as plotted by FlowJo \cite{FlowJo}; \textbf{(B)} Histograms for $P(\log I)$ (red,dots) and $Q(I)$ (blue,dots) vs. $\log I$ as estimated from the data binned logarithmically. Note that plotting $Q(I)$ vs. $\log I$ is somewhat unusual but allows both $P$ and $Q$ to be plotted on the same axis. The red line shows the result of fitting $P(\log I)$ to a gaussian mixture model (Eq.~\ref{eq:LogNormalMix}), and the blue line is the estimate for $Q(I)$ from $P(\log I)$ according to Eq.~\ref{eq:P2Q}. The blue star indicates the location of the only maximum for $Q(I)$ obtained from Eq. \ref{eq:extremaQ}, despite the obvious two maxima in $P(\log I)$ (red). Hartigan's unimodality p-values for $\log I$ (red) and $I$ (blue) are taken directly from the data without binning, corroborating that $\log I$ is bimodal whereas $I$ is unimodal. \textbf{(C)} (Vertical axis represents the functions $A, B_1,B_3$ according to the legend). Graphic solution of the extrema conditions as in Fig.~\ref{fig:ExtremaSolutions} explicitly reveals the three solutions for $P(\log I_*)$ (red line intersects black dashed) as opposed to the single solution for $Q(I_*)$ (the blue line intersects the black dashed line up beyond the plotted area, solution also plotted as blue star in the middle plot), indicating that $Q(I)$ has only one mode.}
  \label{fig:Experiment}
\end{figure}
Given the prevalence and success of manual gating of FCM data in many situations, we wondered whether gating in logarithmic scale could be justified \emph{a posteriori}, based on biological knowledge. In the final part of our analysis, we aimed to include additional information such that the biological significance of the distributions in our single-cell measurements is better captured.

\smallskip
We proceed to analyze experiment 2, these data were taken from Ref.~\cite{Feinerman2008}. We briefly sketch what follows: by considering both ERK1 and ppERK levels, (the two being biologically related and experimentally correlated), we demonstrate that dividing one by the other ameliorates the mismatch between the logarithmic and linear representations. This is done by examining their \emph{joint} probability density function $P_2$; we suggest a probabilistic graphical model that describes the correlations in the system and their dependence on cell-to-cell variability; we suggest a way to normalize for cell-to-cell variability, such that the mismatch (resulting from increased uncertainty) between logarithmic and linear gates is resolved. We then generalize our example and provide a simple experimental and theoretical framework that allows less ambiguous determination of positive/negative gates.

\smallskip
In Fig.~\ref{fig:ppERKvsERK1heatmap}(A), we show the experimental data we will use in our proposed solution. We show a heat map of the \emph{joint} distribution of ppERK ($I_{ppERK}$) and total ERK1 ($I_{ERK1}$) expression in mouse CD8+ T-cells. We estimate their joint distribution by using a kernel density estimator \cite{botev2010}, which is used only for presentation purposes and for demonstrating their conditional independence (Fig.~\ref{S1_Fig}).
Notably, whereas the two modes of ppERK significantly overlap when plotted in the marginal distribution $P(\log I_{ppERK})$, ppERK expression correlates with total ERK1 levels in their joint distribution $P_2(\log I_{ppERK},\log I_{ERK1})$. The relation in Eq.~\ref{eq:P2Q} can be generalized to $P_2(\log I_{ppERK},\log I_{ERK1})$ (see, eg., \cite{Mukhopadhyay2000}), however, it will not be used in this manuscript.
